---
title: "Lab2 Instruction"
author: "Jun Cai"
date: "September 18, 2015"
output:
  pdf_document:
    includes:
      in_header: header.tex
    latex_engine: xelatex
  html_document:
    keep_md: yes
---

### Note

- All R codes are present in boxes with grey background. You can run them in your R Console. Lines leading by `##` are outputs of R codes.

- All R codes are run correctly on my own Mac OS X. When you try them on your own computer, please customize your own working directory.

- The functions in **bold** are recommended to use.

- The Lab Instruction in different formats including .Rmd, .md, .html and .pdf are available on my [GitHub](https://github.com/caijun/DAE/tree/master/Lab2). For your convenience, the data used and results produced in the Lab are also provided in the Github.

Lab2 contains basics for file and directory manipulation, and R data input and output. Rather than a complete collection of functions, I will introduce the frequenty-used functions from my own R experience.

## Part I File and Directory Manipulation

R has a variety of functions for file and directory manipulation. The following are a few examples:

`setwd()` and `getwd()`: used to change or determine the current working directory. It's a good habit to set working directory before your data analysis as all results during your data analysis will be stored in the working directory.

`list.files()` and `list.dirs()`: returns a character vector of names of files or directories under the given directory.

`file.info()`: gives file size, creation time, directory vs. ordinary file status, and so on for each file whose name is in the argument, a character vector.

`file.create()` and `dir.create()`: creates files or directories with the given names if they do not already exist.

`file.exists()` and `dir.exists()`: returns a logical vector indicating whether the given file exists for each name in the first argument, a character vector.

`file.copy()` and `file.rename()`: moves files `from` source path `to` destination path. 

`file.remove()` and **`unlink()`**: deletes the files or directories specified by the first argument, a character vector.

```{r}
# set current working directory to DAE
setwd("/Users/tonytsai/Documents/R/DAE")
getwd()

# list all files including directories under current working directory DAE
list.files()
# extract file information for those files
file.info(list.files())
# list only directories under DAE
list.dirs()
# find all R scripts under DAE and give their full path names (or absolute paths)
list.files(recursive = TRUE, pattern = ".R$", full.names = TRUE)

# create a recursive directory under DAE/Lab2, which stores the TXT data 
# that will be read in Part III.
if(!dir.exists("Lab2/data/CMDSSS/SURF_CLI_CHN_MUL_DAY_V3.0"))
  dir.create("Lab2/data/CMDSSS/SURF_CLI_CHN_MUL_DAY_V3.0", recursive = TRUE)

# create a temporary directory under script
if(!dir.exists("script/tmp")) dir.create("script/tmp")
# create a temporary R script under tmp to say Hello World, Hello R!
file.create("script/tmp/tmp.R")
cat("print('Hello World, Hello R!')", file = "script/tmp/tmp.R")
# excute the R script
source("script/tmp/tmp.R")

# copy tmp.R to helloworld.R
file.copy("script/tmp/tmp.R", "script/helloworld.R")
list.files("script", recursive = TRUE)
# rename helloworld.R to hello.R
file.rename("script/helloworld.R", "script/hello.R")
list.files("script", recursive = TRUE)

# delete all R scripts under script directory except for 20150916.R
# attempt to delete inexistent hellworld.R
file.remove(c("script/hello.R", "script/helloworld.R"))
list.files("script", recursive = TRUE)
# delete the temporary directory that is not empty.
unlink("script/tmp", recursive = TRUE)
list.files("script")
```

To see all the file- and directory-related functions, type the following:

```{r, eval=FALSE}
?files
```

## Part II Capturing R Console Output

R provides functions to save the results that appear in your R Console into a text file.

`sink()`: diverts R output to a file connection.

**`capture.output()`**: sends R output to a character string or file connection.

```{r, eval=FALSE}
# type following codes in your Console
# divert R output to CO2.txt under data directory
sink(file = "data/CO2.txt")
# load R built-in dataset CO2. type ?CO2 to see the decription about CO2 dataset.
data("CO2")
# divert the first 7 rows of CO2
head(CO2, 7)
# end the divertion
sink()
```

```{r}
# sink() does not work in knitr, because it is already used internally to capture 
# results. To make it work, you have to use {} wrap up all aboving code in a 
# single expression and print results to Console explicitly.
{
  sink(file = "data/CO2.txt")
  data("CO2")
  print(head(CO2, 7))
  sink()
}
# the content of CO2.txt will be checked in Part III.
```

```{r}
# generate a vector containing a numeric sequence from 1 to 10
1:10
# send the output to variable x
x <- capture.output(1:10)
x
# divert the first 7 rows of CO2 dataset to CO2.txt
capture.output(head(CO2, 7), file = "data/CO2.txt")
```

## Part III Imports and Exports

`read.table()` and `write.table()`:

`read.csv()` and `write.csv()`:

`read.xls()` and `write.xls()`:

`read.dbf()` and `write.dbf()`:

`load()` and `save()`:

`data()`:

## References

The following are materials on R data import/export that you can access on the Web.

- [R Data Import/Export](https://cran.r-project.org/doc/manuals/r-release/R-data.html)
